---
title: "R Notebook"
output: html_notebook
---
Question to Analyse -

What factors explain the difference in price between an economy ticket and a premium-economy airline ticket?


```{r}
#Load packages

library(statsr)
library(dplyr)
library(ggplot2)

#set working directory

setwd("C:/Users/Sonal Somani/Desktop/IIMInternship/R_code")

#load dataset into R

airlines <- read.csv(paste("SixAirlines.csv",sep=""))

#View dataset
View(airlines)

#Check how your data looks like with their datatypes
str(airlines)
```

```{r}
#To get summary statistics of your data, like mean, median, mode for numeric data etc.

summary(airlines)

```

Since Airline column contains nomimal categorical data, we can draw a bar chart to represent its distribution.

```{r}
#Plotting barchart for the airline frequency distribution
ggplot(airlines, aes(x = AIRLINE, fill = AIRLINE)) + geom_bar()

```

We see that our dataset contains mostly Britist airlines carrier with count around 175.

Let's check which company makes the most aircrafts according to our data.

```{r}
ggplot(airlines, aes(x = AIRCRAFT, fill = AIRCRAFT)) + geom_bar()
```
Clearly Boeing is taking the lead here.

Let's check how the airlines relates to these aircarfts.

```{r}
#Distribution showing aircrafts used by the airlines
table(airlines$AIRLINE, airlines$AIRCRAFT)

```

This shows that we only have data for Boeing and Airbus aircrafts with British and Jet Airlines preferring Boeing Aircrafts.

Since flight duration holds numeric data, let's make a box plot for it and see its distribution.

```{r}

bwplot(Segment ~ income | ownHome, data=seg.df, 
       horizontal=TRUE, xlab="Income")

```

